---
title: "RsendUDP"
format: gfm
---

Send a message in a UDP packet from R

Sends a message in a single UDP packet for debugging and diagnostic purposes.
The original use was to debug parallel processing in R and, using a similar function for C++ OpenMP code (see the package inst directory).

### To install:

	library(devtools)
	devtools::install_github("meekj/RsendUDP")

### Simple Example Usage from R

    library(RsendUDP)
    library(stringr)

    message_string <- str_c(Sys.time(), ' function_name ', Sys.getpid(), ' A diagnostic message\n')
    sendUDP_stat <- sendUDP("127.0.0.1", 1800, message_string)

127.0.0.1 is the localhost IP address, the IP address of another system on your network could also be used
1800 is the port number

sendUDP returns zero if the packet was sent successfully.

Collect messages in a Linux terminal:

    nc -lku 1800

nc on MacOS prints only the first packet, use ncat (from nmap) instead:

    ncat -lku 1800
    
This package should have a listenUDP function.

This version of sendUDP probably does not work on Windows.

### Parallel Processing in R Diagnostic Example

```{r}
#| eval: false

library(future)
library(future.apply)
library(RsendUDP) # Send UDP packets with timing and thread data

## Sample function to run multiple copies in parallel
future_example <- function(sleep_time) {
    sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' future_example ', Sys.getpid(), ' Start\n'))
    Sys.sleep(sleep_time)
    sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' future_example ', Sys.getpid(), ' End\n'))
    return(1L)
}

sleep1      <- 0.050 # 50 milliseconds
num_threads <- 6

## Send the data header
sU <- sendUDP("127.0.0.1", 1800, '\ndate time func pid state\n')

plan(future.mirai::mirai_multisession)

## Dummy data prep - provides a measure of sendUDP overhead time
sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' prep ', Sys.getpid(), ' Start\n'))
sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' prep ', Sys.getpid(), ' End\n'))

## Overall parallel processing time - Start
sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' process ', Sys.getpid(), ' Start\n'))

## Run future_example num_threads times, in parallel
results_list <- future_lapply(1:num_threads,
                              function(thread) { future_example(sleep1) },
                              future.packages = c('RsendUDP'))

## Overall parallel processing time - End
sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' process ', Sys.getpid(), ' End\n'))

## Collect results - post processing
sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' post ', Sys.getpid(), ' Start\n'))
result_sum <- 0
for (thread_result in 1:num_threads) {
    result_sum <- result_sum +  results_list[[thread_result]]
}
sU <- sendUDP("127.0.0.1", 1800, str_c(Sys.time(), ' post ', Sys.getpid(), ' End\n'))

## Send a data complete message, could trigger diagnostic data processing
sU <- sendUDP("127.0.0.1", 1800, 'Done\n')

result_sum # Don't print results in a section that is being timed

```
#### Process the diagnostic / timing results

Copy / paste results collected with "nc -lku 1800" in a terminal:

```{r}
#| message: false
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(forcats)
library(ggplot2)

Title <- 'plan(future.mirai::mirai_multisession)'
threads <- read_table('date time func pid state
2026-02-26 22:55:12.065301 prep 18652 Start
2026-02-26 22:55:12.065988 prep 18652 End
2026-02-26 22:55:12.066391 process 18652 Start
2026-02-26 22:55:12.119132 future_example 19408 Start
2026-02-26 22:55:12.124278 future_example 19364 Start
2026-02-26 22:55:12.126396 future_example 19504 Start
2026-02-26 22:55:12.126561 future_example 19371 Start
2026-02-26 22:55:12.127924 future_example 19397 Start
2026-02-26 22:55:12.130166 future_example 19384 Start
2026-02-26 22:55:12.170069 future_example 19408 End
2026-02-26 22:55:12.17539 future_example 19364 End
2026-02-26 22:55:12.176883 future_example 19504 End
2026-02-26 22:55:12.177115 future_example 19371 End
2026-02-26 22:55:12.178376 future_example 19397 End
2026-02-26 22:55:12.180791 future_example 19384 End
2026-02-26 22:55:12.192683 process 18652 End
2026-02-26 22:55:12.193324 post 18652 Start
2026-02-26 22:55:12.196603 post 18652 End
',col_types = 'cccnc')

theme_jm1 <- theme_bw() + # A decent theme for HTML output
    theme(
        plot.title  = element_text(size = rel(1.5), family = 'Helvetica', face = 'bold'),
        plot.subtitle  = element_text(size = rel(1.3), family = 'Helvetica', face = 'bold'),
        axis.title  = element_text(size = rel(1.5), colour = 'black', face = 'bold'),
        axis.text.x = element_text(angle=0, size = rel(1.5), lineheight = 0.9, colour = 'black', vjust = 1, face = 'bold'),
        axis.text.y = element_text(size = rel(1.5), lineheight = 0.9, colour = 'black', hjust = 1, face = 'bold'),
        strip.text = element_text(size = rel(1.6)),
        legend.title = element_text(size = rel(1.75)),
        legend.key  = element_rect(colour = 'white', fill = 'white'),
        legend.text = element_text(size = rel(1.3))
    )


gen_thread_summary <- function(threads, p_fun_name) { # Thread summary for 'date time' timestamps
    threads <- threads %>% mutate(utime = as.numeric(as.POSIXct(str_c(date,' ',time))),
                                  rtime = utime - min(utime)) %>% select(-date, -time, -utime) %>% arrange(rtime)
    threads <- threads %>% mutate(rtime = 1000 * rtime) # Make it milliseconds
    udp_overhead_time <- max(threads %>% filter(func == 'prep') %>% pull(rtime)) - min(threads %>% filter(func == 'prep') %>% pull(rtime))

    first_thread <- min(threads %>% filter(func == p_fun_name) %>% pull(rtime))- min(threads %>% filter(func == 'process') %>% pull(rtime))
        
    thread_summary <- threads %>% pivot_wider(names_from = state, values_from = rtime) %>% arrange(Start) %>% mutate(dt = End - Start)
    thread_summary <- thread_summary %>% arrange(func, Start)
    thread_summary <- thread_summary %>% filter(func != 'process') %>% select(func, Start, End) %>%
        mutate(func = if_else(func == p_fun_name, str_c(func, '_', row_number()), func))
    thread_summary <- thread_summary %>% arrange(Start) %>% mutate(dt = End - Start)

    ## Threads may not start, or be reported, in order
    func_order          <- thread_summary %>% pull(func) # Retain actual thread start order
    thread_summary$func <- fct_relevel(thread_summary$func, func_order)

    return(list(udp_overhead_time = udp_overhead_time, first_thread = first_thread, thread_summary = thread_summary))
}

plot_r_threads <- function(thread_sum, title) {
    plot_obj <- ggplot(thread_sum, aes(y = func)) +
        geom_segment(aes(x = Start, xend = End, yend = func), linewidth = 4, color = 'blue') +
        labs(x = 'Time, ms', y = '') +
        scale_x_continuous(breaks = seq(0, max(thread_sum$End), 10), minor_breaks = seq(0, max(thread_sum$End), 5)) +
        ggtitle(title) +
        theme_jm1
    return(plot_obj)
}

ts_list <- gen_thread_summary(threads, 'future_example')

Title <- 'plan(future.mirai::mirai_multisession)'
plot_r_threads(ts_list$thread_summary, Title)

ts_list$thread_summary
str_c('UDP Overhead: ', formatC(ts_list$udp_overhead_time, digits = 3, format = 'f'), ' ms   ',
      'First thread: ', formatC(ts_list$first_thread, digits = 3, format = 'f'), ' ms')
                  

```



### Parallel Processing with Rcpp and OpenMP Diagnostic Example


```{r}
#| eval: false

library(Rcpp)

codeFile <- '~/lab/R/omp-udp.cpp'

sourceCpp(rebuild = TRUE, showOutput = TRUE, file=codeFile)

num_threads <- 6
sleep_time  <- 0.05 # 50 ms

## Send a data header
sU <- sendUDP("127.0.0.1", 1800, '\ntime | thread | state\n')

## Send two messages to estimate overhead
sU  <- send_udp_cpp('Zero')
sU  <- send_udp_cpp('Zero')
res <- omp_example(num_threads, sleep_time)

## Result should be count of the number of threads 
res

## Send a data complete message, could trigger diagnostic data processing
sU <- sendUDP("127.0.0.1", 1800, 'Done\n')

```

#### Process the diagnostic / timing results

Raw timestamps are nanoseconds and are converted to milliseconds for analysis.

Copy / paste results collected with "nc -lku 1800" in a terminal:

```{r}

threads <- read_delim('time | thread | state
1772142185910050155 | 0 | Zero
1772142185910315684 | 0 | Zero
1772142185910609342 | 0 | Start
1772142185910687473 | 2 | Start
1772142185910641015 | 1 | Start
1772142185910689335 | 3 | Start
1772142185910692431 | 4 | Start
1772142185910812198 | 5 | Start
1772142185960861013 | 1 | End
1772142185960837190 | 3 | End
1772142185960880301 | 2 | End
1772142185960837174 | 0 | End
1772142185960917800 | 4 | End
1772142185960964650 | 5 | End
', delim = '|', trim_ws = TRUE, show_col_types = FALSE)

gen_thread_summary_ns <- function(threads) { # Thread summary for ns timestamps
    threads <- threads %>% mutate(rtime = (time - min(time)) / 1e6) # rtime in ms
    udp_overhead_time <- max(threads %>% filter(state == 'Zero') %>% pull(rtime)) - min(threads %>% filter(state == 'Zero') %>% pull(rtime))
    threads <- threads %>% mutate(rtime = rtime - max(threads %>% filter(state == 'Zero') %>% pull(rtime)))
    threads <- threads %>% filter(state != 'Zero')

    thread_summary <- threads %>% select(-time) %>% arrange(rtime)
    thread_summary <- thread_summary %>% pivot_wider(names_from = state, values_from = rtime) %>% arrange(Start) %>% mutate(dt = End - Start)
    thread_summary <- thread_summary %>% select(thread, Start, End) %>% mutate(func =  str_c('Thread ', thread))
    thread_summary <- thread_summary %>% arrange(Start)

    return(list(udp_overhead_time = udp_overhead_time, thread_summary = thread_summary))
}

plot_omp_threads <- function(thread_sum, title) {
    func_order <- thread_sum %>% arrange(Start) %>% pull(func)  # Threads may not start, or be reported, in order
    thread_sum$func <- fct_relevel(thread_sum$func, func_order) # Retain actual thread start order

    plot_obj <- ggplot(thread_sum, aes(y = func)) +
        geom_segment(aes(x = Start, xend = End, yend = func), linewidth = 4, color = 'blue') +
        labs(x = 'Time, ms', y = '') +
        scale_x_continuous(breaks = seq(0, max(thread_sum$End), 10), minor_breaks = seq(0, max(thread_sum$End), 5)) +
        ggtitle(title) +
        theme_jm1
    return(plot_obj)
}

ts_list <- gen_thread_summary_ns(threads)

ts_list$udp_overhead_time

ts_list$thread_summary

Title <- 'sendUDP test from C++ OpenMP'
plot_omp_threads(ts_list$thread_summary, Title)

```





