---
title: "RsendUDP"
format: gfm
---

Send a message in a UDP packet from R

Sends a message in a single UDP packet for debugging and diagnostic
purposes.  The original use was to debug parallel processing in R and,
using a similar function for C++ OpenMP code.

A C++ OpenMP example is in the package inst directory. Use
system.file('omp-udp.cpp', package = 'RsendUDP') to get the path to
the example after package installation.

### To install:

	library(devtools)
	devtools::install_github("meekj/RsendUDP")

### Simple Example Usage from R

    library(RsendUDP)
    library(stringr)

    message_string <- str_c(Sys.time(), ' function_name ', Sys.getpid(), ' A diagnostic message\n')
    sendUDP_stat <- sendUDP("127.0.0.1", 1800, message_string)

127.0.0.1 is the localhost IP address, the IP address of another system on your network could also be used
1800 is the port number

sendUDP returns zero if the packet was sent successfully.

Collect messages in a Linux terminal:

    nc -lku 1800

nc on MacOS prints only the first packet, use ncat (from nmap) instead:

    ncat -lku 1800
    
This package should have a listenUDP function.

This version of sendUDP probably does not work on Windows.

### Parallel Processing in R Diagnostic Example

```{r}
#| eval: false

## Claude code + posit-dev-skills assisted
library(mirai)
library(RsendUDP) # Send UDP packets with timing and thread data

## Sample function - mirai tasks get only the named args they need (no closure capture)
parallel_sleep_test <- function(sleep_time, num_threads, ip_address, port_number) {
    sU <- 0
    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, '\ndate time func pid state\n')

    ## Dummy prep - measures sendUDP overhead
    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, str_c(Sys.time(), ' prep ', Sys.getpid(), ' Start\n'))
    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, str_c(Sys.time(), ' prep ', Sys.getpid(), ' End\n'))

    ## Overall parallel start
    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, str_c(Sys.time(), ' process ', Sys.getpid(), ' Start\n'))

    ## Dispatch all tasks first, then collect — minimises time between dispatches
    tasks <- lapply(seq_len(num_threads), function(i) {
        mirai(
            {
                sU <- RsendUDP::sendUDP(ip, port, paste0(Sys.time(), ' future_example ', Sys.getpid(), ' Start\n'))
                Sys.sleep(st)
                sU <- RsendUDP::sendUDP(ip, port, paste0(Sys.time(), ' future_example ', Sys.getpid(), ' End\n'))
                1L
            },
            st   = sleep_time,
            ip   = ip_address,
            port = port_number
        )
    })

    ## Collect results
    results <- lapply(tasks, \(m) m[])

    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, str_c(Sys.time(), ' process ', Sys.getpid(), ' End\n'))

    ## Post processing
    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, str_c(Sys.time(), ' post ', Sys.getpid(), ' Start\n'))
    result_sum <- Reduce(`+`, results, 0L)
    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, str_c(Sys.time(), ' post ', Sys.getpid(), ' End\n'))

    sU <- sU + RsendUDP::sendUDP(ip_address, port_number, 'Done\n')

    return(list(threads = result_sum, UDPres = sU))
}

## First start the UDP packet collection code in process-udp.R

num_threads <- 6

## Start persistent daemons, pre-loading packages so they are NOT re-exported per call.
## dispatcher = FALSE is lower latency when you always submit exactly num_threads tasks.
daemons(num_threads, dispatcher = FALSE)

## Pre-load packages on all workers once via broadcast — not re-exported per task.
everywhere(library(RsendUDP, quietly = TRUE))

sleep1 <- 0.050 # 50 milliseconds

parallel_sleep_test(sleep1, num_threads, '127.0.0.1', 1800)

## Clean up daemons when done
## daemons(0)

```
#### Process the diagnostic / timing results

See process-udp.R in the inst directory, or
copy / paste results collected with "nc -lku 1800" in a terminal:

```{r}
#| message: false
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(forcats)
library(ggplot2)

Title <- '{mirai} Sleep Test Diagnostic'
threads <- read_table('date time func pid state
2026-04-04 01:59:23.640762 prep 659577 Start
2026-04-04 01:59:23.641291 prep 659577 End
2026-04-04 01:59:23.641563 process 659577 Start
2026-04-04 01:59:23.6423 future_example 3075982 Start
2026-04-04 01:59:23.642483 future_example 3075997 Start
2026-04-04 01:59:23.64262 future_example 3076073 Start
2026-04-04 01:59:23.642683 future_example 3076120 Start
2026-04-04 01:59:23.642817 future_example 3076137 Start
2026-04-04 01:59:23.643005 future_example 3076183 Start
2026-04-04 01:59:23.692602 future_example 3075982 End
2026-04-04 01:59:23.692729 future_example 3075997 End
2026-04-04 01:59:23.692911 future_example 3076120 End
2026-04-04 01:59:23.692889 future_example 3076073 End
2026-04-04 01:59:23.693055 future_example 3076137 End
2026-04-04 01:59:23.693246 future_example 3076183 End
2026-04-04 01:59:23.6935 process 659577 End
2026-04-04 01:59:23.694066 post 659577 Start
2026-04-04 01:59:23.694514 post 659577 End
',col_types = 'cccnc')

theme_jm1 <- theme_bw() + # A decent theme for HTML output
    theme(
        plot.title  = element_text(size = rel(1.5), family = 'Helvetica', face = 'bold'),
        plot.subtitle  = element_text(size = rel(1.3), family = 'Helvetica', face = 'bold'),
        axis.title  = element_text(size = rel(1.5), colour = 'black', face = 'bold'),
        axis.text.x = element_text(angle=0, size = rel(1.5), lineheight = 0.9, colour = 'black', vjust = 1, face = 'bold'),
        axis.text.y = element_text(size = rel(1.5), lineheight = 0.9, colour = 'black', hjust = 1, face = 'bold'),
        strip.text = element_text(size = rel(1.6)),
        legend.title = element_text(size = rel(1.75)),
        legend.key  = element_rect(colour = 'white', fill = 'white'),
        legend.text = element_text(size = rel(1.3))
    )


gen_thread_summary <- function(threads, p_fun_name) { # Thread summary for 'date time' timestamps
    threads <- threads %>% mutate(utime = as.numeric(as.POSIXct(str_c(date,' ',time))),
                                  rtime = utime - min(utime)) %>% select(-date, -time, -utime) %>% arrange(rtime)
    threads <- threads %>% mutate(rtime = 1000 * rtime) # Make it milliseconds
    udp_overhead_time <- max(threads %>% filter(func == 'prep') %>% pull(rtime)) - min(threads %>% filter(func == 'prep') %>% pull(rtime))

    first_thread <- min(threads %>% filter(func == p_fun_name) %>% pull(rtime))- min(threads %>% filter(func == 'process') %>% pull(rtime))
        
    thread_summary <- threads %>% pivot_wider(names_from = state, values_from = rtime) %>% arrange(Start) %>% mutate(dt = End - Start)
    thread_summary <- thread_summary %>% arrange(func, Start)
    thread_summary <- thread_summary %>% filter(func != 'process') %>% select(func, Start, End) %>%
        mutate(func = if_else(func == p_fun_name, str_c(func, '_', row_number()), func))
    thread_summary <- thread_summary %>% arrange(Start) %>% mutate(dt = End - Start)

    ## Threads may not start, or be reported, in order
    func_order          <- thread_summary %>% pull(func) # Retain actual thread start order
    thread_summary$func <- fct_relevel(thread_summary$func, func_order)

    return(list(udp_overhead_time = udp_overhead_time, first_thread = first_thread, thread_summary = thread_summary))
}

plot_r_threads <- function(thread_sum, title) {
    plot_obj <- ggplot(thread_sum, aes(y = func)) +
        geom_segment(aes(x = Start, xend = End, yend = func), linewidth = 4, color = 'blue') +
        labs(x = 'Time, ms', y = '') +
        scale_x_continuous(breaks = seq(0, max(thread_sum$End), 10), minor_breaks = seq(0, max(thread_sum$End), 5)) +
        ggtitle(title) +
        theme_jm1
    return(plot_obj)
}

ts_list <- gen_thread_summary(threads, 'future_example')

plot_r_threads(ts_list$thread_summary, Title)

ts_list$thread_summary
str_c('UDP Overhead: ', formatC(ts_list$udp_overhead_time, digits = 3, format = 'f'), ' ms   ',
      'First thread: ', formatC(ts_list$first_thread, digits = 3, format = 'f'), ' ms')
                  

```



### Parallel Processing with Rcpp and OpenMP Diagnostic Example


```{r}
#| eval: false

library(Rcpp)

codeFile <- system.file('omp-udp.cpp', package = 'RsendUDP')
sourceCpp(rebuild = TRUE, showOutput = TRUE, file=codeFile)

num_threads <- 6
sleep_time  <- 0.05 # 50 ms

## Send a data header
sU <- sendUDP("127.0.0.1", 1800, '\ntime | thread | state\n')

## Send two messages to estimate overhead
sU  <- send_udp_cpp('Zero')
sU  <- send_udp_cpp('Zero')
res <- omp_example(num_threads, sleep_time)

## Result should be count of the number of threads 
res

## Send a data complete message, could trigger diagnostic data processing
sU <- sendUDP("127.0.0.1", 1800, 'Done\n')

```

#### Process the diagnostic / timing results

Raw timestamps are nanoseconds and are converted to milliseconds for analysis.

Copy / paste results collected with "nc -lku 1800" in a terminal:

```{r}

threads <- read_delim('time | thread | state
1772142185910050155 | 0 | Zero
1772142185910315684 | 0 | Zero
1772142185910609342 | 0 | Start
1772142185910687473 | 2 | Start
1772142185910641015 | 1 | Start
1772142185910689335 | 3 | Start
1772142185910692431 | 4 | Start
1772142185910812198 | 5 | Start
1772142185960861013 | 1 | End
1772142185960837190 | 3 | End
1772142185960880301 | 2 | End
1772142185960837174 | 0 | End
1772142185960917800 | 4 | End
1772142185960964650 | 5 | End
', delim = '|', trim_ws = TRUE, show_col_types = FALSE)

gen_thread_summary_ns <- function(threads) { # Thread summary for ns timestamps
    threads <- threads %>% mutate(rtime = (time - min(time)) / 1e6) # rtime in ms
    udp_overhead_time <- max(threads %>% filter(state == 'Zero') %>% pull(rtime)) - min(threads %>% filter(state == 'Zero') %>% pull(rtime))
    threads <- threads %>% mutate(rtime = rtime - max(threads %>% filter(state == 'Zero') %>% pull(rtime)))
    threads <- threads %>% filter(state != 'Zero')

    thread_summary <- threads %>% select(-time) %>% arrange(rtime)
    thread_summary <- thread_summary %>% pivot_wider(names_from = state, values_from = rtime) %>% arrange(Start) %>% mutate(dt = End - Start)
    thread_summary <- thread_summary %>% select(thread, Start, End) %>% mutate(func =  str_c('Thread ', thread))
    thread_summary <- thread_summary %>% arrange(Start)

    return(list(udp_overhead_time = udp_overhead_time, thread_summary = thread_summary))
}

plot_omp_threads <- function(thread_sum, title) {
    func_order <- thread_sum %>% arrange(Start) %>% pull(func)  # Threads may not start, or be reported, in order
    thread_sum$func <- fct_relevel(thread_sum$func, func_order) # Retain actual thread start order

    plot_obj <- ggplot(thread_sum, aes(y = func)) +
        geom_segment(aes(x = Start, xend = End, yend = func), linewidth = 4, color = 'blue') +
        labs(x = 'Time, ms', y = '') +
        scale_x_continuous(breaks = seq(0, max(thread_sum$End), 10), minor_breaks = seq(0, max(thread_sum$End), 5)) +
        ggtitle(title) +
        theme_jm1
    return(plot_obj)
}

ts_list <- gen_thread_summary_ns(threads)

ts_list$udp_overhead_time

ts_list$thread_summary

Title <- 'sendUDP test from C++ OpenMP'
plot_omp_threads(ts_list$thread_summary, Title)

```





